Bifurcation or state tipping: assessing transition type in a model trophic cascade

### 2018-07-15 paper/paper.Rmd
remove confusing previous sentence at end of intro

Central to resolving this question is untangling a second issue that frequently arises in the discussion of alternative stable state dynamics in ecological systems: the issue of dimension.  The S-tipping and B-tipping transitions illustrated in Figure \ref{fig1} are fundamentally rooted in a one-dimensional analysis: the state consists of a single continuously valued dynamical variable (i.e. population density of adult bass).  Ecosystems are notoriously more complex, and this lake ecosystem is no exception. Here, we analyze a previously published model of a trophic cascade to determine whether manipulating the top predator constitutes B-tipping or S-tipping. Common versions of this model include 5 state variables, ranging from top predators to primary producers (Figure \ref{fig2}). We perform dimension reduction on the full 5D model, and surmise whether the reduced model can be expressed in a way such that manipulating predator abundance is tantamount to manipulating a parameter, or if it necessarily constitutes state manipulation. This exercise facilitates the interpretation of the lake experiment in the context of predicting regime shifts by using a 1D model to understand a high dimensional system.

>>>>

Central to resolving this question is untangling a second issue that frequently arises in the discussion of alternative stable state dynamics in ecological systems: the issue of dimension.  The S-tipping and B-tipping transitions illustrated in Figure \ref{fig1} are fundamentally rooted in a one-dimensional analysis: the state consists of a single continuously valued dynamical variable (i.e. population density of adult bass).  Ecosystems are notoriously more complex, and this lake ecosystem is no exception. Here, we analyze a previously published model of a trophic cascade to determine whether manipulating the top predator constitutes B-tipping or S-tipping. Common versions of this model include 5 state variables, ranging from top predators to primary producers (Figure \ref{fig2}). We perform dimension reduction on the full 5D model. This exercise facilitates the interpretation of the lake experiment in the context of predicting regime shifts by using a 1D model to understand a high dimensional system.
